On centrality of even orthogonal $\Kt$

### 2016-06-01 paper.tex
Andrei's fixes

\documentclass[11pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, verbatim, enumitem, mathtools, xfrac, nameref, thmtools, hyperref}
\usepackage[utf8]{inputenc}
\usepackage[backend=biber, citestyle=numeric-comp, natbib=true, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}            
\usepackage[activate={true,nocompatibility}]{microtype}
\usepackage[capitalize]{cleveref}
\usepackage[english]{babel} 
\usepackage{stmaryrd}
\usepackage[matrix,arrow,curve]{xy}

@@

Recently the authors of this paper have established the centrality of $\Kt(\rC_\ell, R)$ ($\ell\geq 3$) and $\Kt(\rE_\ell, R)$ ($\ell=6,7,8$) for an arbitrary commutative ring $R$, see \cite{Lav, SCh}.
In turn, the key result of this paper is the following theorem.
\begin{tm}[``Centrality of $\Kt$''] \label{tm:centrality}  Let $R$ be an arbitrary commutative ring.
Then for $\ell\geq 4$ the group $\Kt(\rD_\ell, R)$ is a central subgroup of $\St(\rD_\ell, R)$. \end{tm}

@@

\section{Preliminaries} \label{sec:prelim}
Throughout this paper $R$ denotes an associative commutative ring with identity.
All commutators are left-normed, i.\,e. $[x,\,y]=xyx\inv y\inv$. 
We denote by $R^n$ the free right $R$-module with a basis $e_1,\ldots,e_n$ and by $\Um(n,\,R)$ the subset of unimodular columns $v\in R^n$, i.\,e. 
columns whose components generate $R$ as an ideal.

First of all, recall that for a simply laced root system $\Phi$ of rank $\geq 2$ the Steinberg group $\St(\Phi, R)$ 
is defined by generators $x_\alpha(r)$, $\alpha\in\Phi$, $r\in R$ and the following relations:
\begin{align}
 x_\alpha(r) x_\alpha(s) & = x_\alpha(r+s), & \nonumber \\
 [x_\alpha(r),  x_\beta(s)] & = x_{\alpha+\beta}(N_{\alpha,\beta}\cdot rs), &\ \alpha,\beta\in \Phi,\ \alpha+\beta\in\Phi, \nonumber \\
 [x_\alpha(r),  x_\beta(s)] & = 1, &\ \alpha,\beta\in \Phi,\ \alpha+\beta\not\in\Phi. \nonumber
\end{align}
Here $N_{\alpha,\beta}$ are the structure constants of $\Phi$ which are equal to $\pm 1$.
We use more a more familiar notation for root unipotents in the case $\Phi=\rA_\ell$:
$x_{ij}(r),\ 1\leq i\neq j\leq n,\ r\in R,$
\setcounter{equation}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\begin{align}
x_{ij}(r)x_{ij}(s)      & = x_{ij}(r+s),& \label{add0}\\
[x_{ij}(r),\,x_{hk}(s)] & = 1,& \text{ for }h\neq j,\ k\neq i, \label{ccf1}\\
[x_{ij}(r),\,x_{jk}(s)] & = x_{ik}(rs).& \label{ccf2}
\end{align}
In this situation the elementary root unipotents coincide with elementary transvections $t_{ij}(r)=1+r \cdot e_{ij}$, $1\leq i\neq j\leq n$, $r\in R$
and the elementary subgroup $\E(\rA_{n-1}, R)$ is denoted by $\E(n, R)$.
Here $1$ denotes the identity matrix and $e_{ij}$ ientriess the standard matrix unit. 
The natural projection $\phi\colon\St(n,\,R)\rightarrow\E(n,\,R)$ sends $x_{ij}(r)$ to $t_{ij}(r)$.

In \cite{Sus} Suslin showed that for $n\geq 3$ the elementary group $\E(n, R)$ coincides with the subgroup of $\GL(n,\,R)$ generated by matrices
of the form $t(u,\,v)=1+uv^t$ where $u\in\Um(n,\,R)$, $v\in R^n$ and $u^tv=0$. Here $u^t$ stands for the transpose of $u$.
This result clearly implies that $\E(n,\,R)$ is normal in $\GL(n,\,R)$.
Developing Suslin's ideas, W.~van~der~Kallen showed that for $n\geq4$ the Steinberg group $\St(n,\,R)$ is isomorphic to the group with the following presentation by generators
$$\{X(u,\,v)\mid u\in\Um(n,\,R),\ v\in R^n,\ u^tv=0\}$$ and relations:
\setcounter{equation}{0} \renewcommand{\theequation}{K\arabic{equation}}
\begin{align}
&X(u,\,v)X(u,\,w)=X(u,\,v+w), \label{add1} \\
&X(u,\,v)X(u',\,v')X(u,\,v)\inv=X(t(u,\,v)u',\,t(v,\,u)\inv v'). \label{conj1}
\end{align}
This presentation clearly implies that $\phi\colon\St(n,\,R)\rightarrow\E(n,\,R)$ is a central extension.
Notice that we parametrize elements $X(u, v)$ by two columns rather than by a column and a row as it is done in~\cite{vdK}.
Our generator $X(u,v)$ corresponds to van der Kallen's generator $X(u, v^t)$.

@@

In the case of an arbitrary $I$ the relative Steinberg group $\St(\Phi, R, I)$ does not have to be a subgroup of $\St(\Phi, R)$ and should be defined as a central extension of the above kernel.
For the purposes of the present text it suffices to consider only relative Steinberg groups corresponding to splitting ideals.
For more information regarding the general case we refer the reader to \cite[Section~3]{SCh}.

@@

We claim that the main results of this paper follow from Tulenbaev's lifting property.
\begin{prop} For any irreducible $\Phi$ of rank $\geq 2$ there are implications $a) \implies b) \implies c) \implies d)$.
\begin{enumerate}
 \item the Steinberg group functor $\St(\Phi, -)$ satisfies lifting property~\eqref{eq:diagram};
 \item a dilation principle holds for $\St(\Phi, -)$, i.\,e. if $g\in\St(\Phi, A[X], XA[X])$ is such that $\lambda_a^*(g) = 1 \in \St(\Phi, R_a[X])$ then
       for sufficiently large $n$ one has $ev_{X = a^n Y}^*(h) = 1 \in \St(\Phi, A[Y])$, where $ev_{X=a^n Y}$ denotes the unique $R$-algebra morphism $R[X]\to R[Y]$ sending $X$ to $a^nY$;
 \item the assertion of \cref{tm:lg-principle} holds for $\St(\Phi, -)$;
 \item $\Kt(\Phi, R)$ is a central subgroup of $\St(\Phi, R)$.
\end{enumerate} \end{prop}
\begin{proof} The specified implications are contained in the proofs of Lemma~15, Theorem~2 and Theorem~1 of \cite{SCh}, respectively. \end{proof}

@@

Now denote by $G_0$ the free product of Steinberg groups $\St(\Psi, R, I)$ where $\Psi$ varies over all root subsystems of $\Phi$ which have type $\rA_3$.
Denote by $z_\alpha^\Psi(s,r)$ the image of $z_\alpha(s,r)\in\St(\Psi, R, I)$ under the canonical embedding $\St(\Psi, R, I) \hookrightarrow G_0$.
Now let $G$ be the quotient of $G_0$ modulo all relations of the form $z_\alpha^{\Psi_1}(s, r) = z_\alpha^{\Psi_2}(s, r)$, where $\alpha\in\Psi_1\cap\Psi_2$, $s\in I$, $r\in I$ 
and both $\Psi_1$ and $\Psi_2$ have type $\rA_3$.
The following result has been proved in~\cite{SCh}, see~Theorem~9 combined with Remark~3.16.
\begin{tm*} \label{tm:relPres} For a simply laced irreducible root system $\Phi$ of rank $\geq 3$ the canonical map $G \rightarrow \St(\Phi, R, I)$ is an isomorphism of groups.
\end{tm*}

@@

As for the map $\iota$, we first define it as a homomorphism to the absolute group. % which does not rely on the assumption that $I$ is splitting.
Then we prove that the image of $\iota$ is contained in $\St(n, R, I)$.
%$$\xymatrix{\St^*(n,\,R,\,I)\ar@<0.0ex>[r]^{\iota}&\St(n,\,R).}$$
Clearly, $\iota$ the elements $F(u,\,v)$ should go to van der Kallen's generators $X(u,\,v)$.
But now we should find the images for the elements $S(u,\,v)$ as well.
First of all, recall that van der Kallen in~\cite[3.8--3.10]{vdK} constructs the following elements:
$$x(u,\,v)\in\St(n,\,R),\ u^tv=0,\ u_i=0\,\text{ or }\,v_i=0\,\text{ for some }\,1\leq i\leq n.$$
The following definition is the ``transpose'' of \cite[3.13]{vdK}.
\begin{df} For $u\in R^n$, $v\in\E(n,\,R)e_1$, $u^tv=0$, consider the set $\overline Y(u,\,v)\subseteq\St(n,\,R)$
 consisiting of all elements $y\in\St(n,\,R)$ that admit a decomposition $y=\prod x(u^k,\,v)$, 
 where $\sum u^k=u$ and $u^k= (e_pv_q-e_qv_p)c_k$ for some $c_k\in R$, $1\leq p\neq q\leq n$. \end{df}

@@

\begin{prop} \label{lm:map-iota}
 For a splitting ideal $I \trianglelefteq R$ and $n\geq 4$ the groups $\St^*(n,\,R,\,I)$ and $\St(n,\,R, I)$ are isomorphic.
\end{prop}

@@

We will also need the ``transposed'' analogue of Tulenbaev's elements $X_{u,v}(a)$.
\begin{df} \label{df:TulY}
Let $v\in B^n$, $a\in I(v)$, and $u \in D(v)$, i.\,e. $u = \sum\limits_{i=1}^M u_i$ for some $u_k$ such that $v^tu_k=0$ and each $u_k$ has two zero components.
Set $Y_{u,v}(a)=\prod_{k=1}^Mx(u_ka,\,v)$. 
Similarly to \ref{df:TulX} one can check that this definition does not depend either on the order of factors or on the choice of decomposition for $u$.
\end{df}

>>>>

\documentclass[11pt]{amsart}
\usepackage{amscd, amsmath, amssymb, amsthm, amsfonts, amstext, verbatim, enumitem, mathtools, xfrac, nameref, thmtools, hyperref}
\usepackage[utf8]{inputenc}
\usepackage[backend=bibtex, citestyle=numeric-comp, natbib=true, sortlocale=en_US, url=false, doi=false, eprint=true, maxbibnames=4]{biblatex}            
\usepackage[activate={true,nocompatibility}]{microtype}
\usepackage[capitalize]{cleveref}
\usepackage[english]{babel} 
\usepackage{stmaryrd}
\usepackage[matrix,arrow,curve]{xy}

@@

Recently the authors of this paper have established the centrality of $\Kt(\rC_\ell, R)$ ($\ell\geq 3$) and $\Kt(\rE_\ell, R)$ ($\ell=6,7,8$) for an arbitrary commutative ring $R$, see \cite{Lav, SCh}.
In turn, the key result of this paper is the following theorem.
\begin{tm}[``Centrality of $\Kt$''] \label{tm:centrality}  Let $R$ be an arbitrary commutative ring.
Then, for $\ell\geq 4$ the group $\Kt(\rD_\ell, R)$ is a central subgroup of $\St(\rD_\ell, R)$. \end{tm}

@@

\section{Preliminaries} \label{sec:prelim}
Throughout this paper $R$ denotes an associative commutative ring with identity.
All commutators are left-normed, i.\,e. $[x,\,y]=xyx\inv y\inv$. 
We denote by $R^n$ the free right $R$-module with a basis $e_1,\ldots,e_n$ and by $\Um(n,\,R)$ the set of unimodular columns $v\in R^n$, i.\,e. 
columns whose components generate $R$ as an ideal.

First of all, recall that for a simply laced root system $\Phi$ of rank $\geq 2$ the Steinberg group $\St(\Phi, R)$ 
is defined by generators $x_\alpha(r)$, $\alpha\in\Phi$, $r\in R$ and the following relations:
\begin{align}
 x_\alpha(r) x_\alpha(s) & = x_\alpha(r+s), & \nonumber \\
 [x_\alpha(r),  x_\beta(s)] & = x_{\alpha+\beta}(N_{\alpha,\beta}\cdot rs), &\ \alpha,\beta\in \Phi,\ \alpha+\beta\in\Phi, \nonumber \\
 [x_\alpha(r),  x_\beta(s)] & = 1, &\ \alpha,\beta\in \Phi,\ \alpha+\beta\not\in\Phi. \nonumber
\end{align}
Here $N_{\alpha,\beta}$ are the structure constants of $\Phi$ which are equal to $\pm 1$.
We use a more familiar notation for root unipotents in the case $\Phi=\rA_\ell$:
$x_{ij}(r),\ 1\leq i\neq j\leq n,\ r\in R,$
\setcounter{equation}{0}
\renewcommand{\theequation}{S\arabic{equation}}
\begin{align}
x_{ij}(r)x_{ij}(s)      & = x_{ij}(r+s),& \label{add0}\\
[x_{ij}(r),\,x_{hk}(s)] & = 1,& \text{ for }h\neq j,\ k\neq i, \label{ccf1}\\
[x_{ij}(r),\,x_{jk}(s)] & = x_{ik}(rs).& \label{ccf2}
\end{align}
In this situation the elementary root unipotents coincide with elementary transvections $t_{ij}(r)=1+r \cdot e_{ij}$, $1\leq i\neq j\leq n$, $r\in R$
and the elementary subgroup $\E(\rA_{n-1}, R)$ is denoted by $\E(n, R)$.
Here $1$ denotes the identity matrix and $e_{ij}$ is the standard matrix unit. 
The natural projection $\phi\colon\St(n,\,R)\rightarrow\E(n,\,R)$ sends $x_{ij}(r)$ to $t_{ij}(r)$.

In \cite{Sus} Suslin showed that for $n\geq 3$ the elementary group $\E(n, R)$ coincides with the subgroup of $\GL(n,\,R)$ generated by matrices
of the form $t(u,\,v)=1+uv^t$ where $u\in\Um(n,\,R)$, $v\in R^n$ and $u^tv=0$. Here $u^t$ stands for the transpose of $u$.
This result clearly implies that $\E(n,\,R)$ is normal in $\GL(n,\,R)$.
Developing Suslin's ideas, van~der~Kallen showed that for $n\geq4$ the Steinberg group $\St(n,\,R)$ is isomorphic to the group with the following presentation by generators
$$\{X(u,\,v)\mid u\in\Um(n,\,R),\ v\in R^n,\ u^tv=0\}$$ and relations:
\setcounter{equation}{0} \renewcommand{\theequation}{K\arabic{equation}}
\begin{align}
&X(u,\,v)X(u,\,w)=X(u,\,v+w), \label{add1} \\
&X(u,\,v)X(u',\,v')X(u,\,v)\inv=X(t(u,\,v)u',\,t(v,\,u)\inv v'). \label{conj1}
\end{align}
This presentation clearly implies that $\phi\colon\St(n,\,R)\rightarrow\E(n,\,R)$ is a central extension.
Notice that we parametrize elements $X(u, v)$ by two columns rather than by a column and a row as it is done in~\cite{vdK}.
Our generator $X(u,v)$ corresponds to van der Kallen's generator $X(u, v^t)$.

@@

In the case of an arbitrary $I$ the relative Steinberg group $\St(\Phi, R, I)$ is not a subgroup of $\St(\Phi, R)$ and should be defined as a central extension of the above kernel.
For the purposes of the present text it suffices to consider only relative Steinberg groups corresponding to splitting ideals.
For more information regarding the general case we refer the reader to \cite[Section~3]{SCh}.

@@

We claim that the main results of this paper follow from Tulenbaev's lifting property.
\begin{prop} For any irreducible $\Phi$ of rank $\geq 2$ hold implications $a) \implies b) \implies c) \implies d)$.
\begin{enumerate}
 \item the Steinberg group functor $\St(\Phi, -)$ satisfies lifting property~\eqref{eq:diagram};
 \item a dilation principle holds for $\St(\Phi, -)$, i.\,e. if $g\in\St(\Phi, A[X], XA[X])$ is such that $\lambda_a^*(g) = 1 \in \St(\Phi, R_a[X])$ then
       for a sufficiently large $n$ one has $ev_{X = a^n Y}^*(h) = 1 \in \St(\Phi, A[Y])$, where $ev_{X=a^n Y}$ denotes the unique $R$-algebra morphism $R[X]\to R[Y]$ sending $X$ to $a^nY$;
 \item the assertion of \cref{tm:lg-principle} holds for $\St(\Phi, -)$;
 \item $\Kt(\Phi, R)$ is a central subgroup of $\St(\Phi, R)$.
\end{enumerate} \end{prop}
\begin{proof} The specified implications are contained in the proofs of Lemma~15, Theorem~2 and Theorem~1 of \cite{SCh}, respectively. \end{proof}

@@

Now, denote by $G_0$ the free product of Steinberg groups $\St(\Psi, R, I)$ where $\Psi$ varies over all root subsystems of $\Phi$ which have type $\rA_3$.
Denote by $z_\alpha^\Psi(s,r)$ the image of $z_\alpha(s,r)\in\St(\Psi, R, I)$ under the canonical embedding $\St(\Psi, R, I) \hookrightarrow G_0$.
Now let $G$ be the quotient of $G_0$ modulo all relations of the form $z_\alpha^{\Psi_1}(s, r) = z_\alpha^{\Psi_2}(s, r)$, where $\alpha\in\Psi_1\cap\Psi_2$, $s\in I$, $r\in I$ 
and both $\Psi_1$ and $\Psi_2$ have type $\rA_3$.
The following result has been proved in~\cite{SCh}, see~Theorem~9 combined with Remark~3.16.
\begin{tm*} \label{tm:relPres} For a simply laced irreducible root system $\Phi$ of rank $\geq 3$ the canonical map $G \rightarrow \St(\Phi, R, I)$ is an isomorphism of groups.
\end{tm*}

@@

As for the map $\iota$, we first define it as a homomorphism to the absolute group. % which does not rely on the assumption that $I$ is splitting.
Then we prove that the image of $\iota$ is contained in $\St(n, R, I)$.
%$$\xymatrix{\St^*(n,\,R,\,I)\ar@<0.0ex>[r]^{\iota}&\St(n,\,R).}$$
Clearly, the elements $F(u,\,v)$ should go to van der Kallen's generators $X(u,\,v)$.
But now we should find the images for the elements $S(u,\,v)$ as well.
First of all, recall that van der Kallen in~\cite[3.8--3.10]{vdK} constructs the following elements:
$$x(u,\,v)\in\St(n,\,R),\ u^tv=0,\ u_i=0\,\text{ or }\,v_i=0\,\text{ for some }\,1\leq i\leq n.$$
The following definition is the ``transpose'' of \cite[3.13]{vdK}.
\begin{df} For $u\in R^n$, $v\in\E(n,\,R)e_1$, $u^tv=0$, consider the set $\overline Y(u,\,v)\subseteq\St(n,\,R)$
 consisiting of all elements $y\in\St(n,\,R)$ that admit a decomposition $y=\prod x(u^k,\,v)$, 
 where $\sum u^k=u$ and $u^k= (e_pv_q-e_qv_p)c_k$ for some $c_k\in R$, $1\leq p\neq q\leq n$. \end{df}

@@

\begin{prop} \label{lm:map-iota}
 For a splitting ideal $I \trianglelefteq R$ and $n\geq 4$ the groups $\St^*(n,\,R,\,I)$ and $\St(n,\,R,\,I)$ are isomorphic.
\end{prop}

@@

We will also need the ``transposed'' analogue of Tulenbaev's elements $X_{u,v}(a)$.
\begin{df} \label{df:TulY}
Let $v\in B^n$, $a\in I(v)$, and $u \in D(v)$, i.\,e. $u = \sum\limits_{i=1}^M u_i$ for some $u_k$ such that $v^tu_k=0$ and each $u_k$ has two zero components.
Set $Y_{u,v}(a)=\prod_{k=1}^Mx(u_ka,\,v)$. 
Similarly to \ref{df:TulX} one can check that this definition does depend neither on the order of factors nor on the choice of decomposition for $u$.
\end{df}
